---
title: "A Tutorial to Genome Explorer in R"
author: "Kari Lavikka"
date: "Mar 16, 2017"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is a short tutorial to Genome Explorer, a JavaScipt library and R package for visualizing genomic coordinates. It is most conveniently used with RStudio and/or Shiny, but it also works with vanilla R.

## Installation

The package is available in Anduril R repository. Installation is easy:

```{r eval=FALSE}
install.packages("GenomeExplorer", repos = "http://anduril.org/pub/R")
```

Or if you want to live on the bleeding edge:

```{r eval=FALSE}
devtools::install_github("tuner/genome-explorer")
```

## Our first visualization

Let's first import the library. We also use magrittr to conveniently add features to Genome Explorer by using the pipe **%>%** operator.

```{r}
library(GenomeExplorer)
library(magrittr)
```

Here is an example with the minimum amount of data and no extra options of any kind. The data contains a single genomic region, which we visualize using the default *simple* visualization. Genome Explorer comes bundled with chromosome data (cytobands) for a few common genome assemblies. Here we use *hg38*.

By default, Genome Explorer expects to find chromosome, start coordinate, and end coordinate from _chrom_, _start_, and _end_ columns, respectively.

```{r}
data <- data.frame(
  chrom = c("chr5"),
  start = c(50000000),
  end = c(150000000)
)

GenomeExplorer("hg38") %>%
  segmentTrack(data)
```

## A more advanced example

```{r echo = FALSE, message = FALSE}
data <- read.table("example_data.txt", header = TRUE, sep = "\t")
```

Our data now contais two new columns: _patient_ and _type_.

```{r echo = FALSE}
data
```

Genome Explorer can display several tracks concurrently, and each track may have a different dataset and visualization type. We will stick with just one track for now.

As we would like to compare the patients with one another, we use the _patient_ column as a discriminator. Each patient then gets an own subtrack. _Type_ is just an arbitrary column that we want to use for coloring the bars produced by the _simple_ visualization.

```{r}
GenomeExplorer("hg38") %>%
  segmentTrack(
    data,
    discriminator = "patient",
    vis = "simple",
    vis_config = list(
      color = "type"
    )
  )
```

## A complete example

```{r echo = FALSE, message = FALSE}
data <- read.table("example_cnv.tsv", header = TRUE, sep = "\t")
```

We finally move on to a real use case. Our data now contains genomic ranges with CNV (Copy number variation) and BAF (B Allele Frequency) values. Here's a sample of the dataset:

```{r echo = FALSE}
data[1:8, ]
```

The columns that define the genomic coordinates are now named differently. Fortunately, you do not need to rename the columns - just define them explicitly by attributes _chrom_, _start_, and _end_.

Because we would like to compare the diagnosis (D) and relapse (R) samples side by side, we explicitly enumerate the cases in a specific order. By default, they are ordered alphanumerically.

Genome Explorer provides you with a particular visualization (_cnvAndBaf_) for CNV and BAF data. CNV, which is expressed as *segment mean*, is displayed as a color, ranging from blue (negative, deletion) to red (positive, amplification) through white (zero, no change). BAF is actually displayed as LOH (Loss of Heterozygosity), which is calculated using the formula: |*baf* - 0.5| * 2. The height of the bar is determined by LOH, 0% means full heterozygosity, 100% mean total loss of the B allele.

We also define a title for the data track. Looks prettier that way.

As a final touch we include a gene annotation track, which gives our data some context. The genes and transcripts become visible when you zoom in enough.

```{r}
GenomeExplorer("hg38") %>%
  segmentTrack(
    data,
    
    # Define the columns for genomic coordinates
    chrom = "Chr",
    start = "Start",
    end = "End",
    
    discriminator = "case",
    
    samples = c("D1", "R1",
                "D2", "R2"),
    
    vis = "cnvAndBaf",
    vis_config = list(
      # Define the columns. Defining both of them is not mandatory.
      cnv = "segMean",
      baf = "bafMean"
    ),
    
    title = "CNV & LOH"
  ) %>%
  refseqGeneTrack()
```

## Best practices

TODO
